SI: Settlements

### 2022-06-23 analysis/paper/paper.Rmd
update whole paper fix lua filters

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = FALSE,
  comment = "#>",
  # fig.path = "../figures/",
  dpi = 300
)
```

@@

```{r get-data, cache = FALSE}
derived_data <- "analysis/data/derived_data"
path_maps <- here::here(derived_data, "maps/")
path_data <- here::here(derived_data)
path_temp <- here::here(derived_data, "temp/")

@@

# labels
labs_chrono <- read_rds(here::here(derived_data, "chrono_labels.RDS"))

# database
set_base <- read_rds(here::here(derived_data, "settlements.RDS"))

set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"), 
                    quiet = TRUE)

@@

<!-- ```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."} -->
<!-- rm_dist %>% -->
<!--   ggplot(aes(x = min, y = rm, fill = region)) + -->
<!--   geom_violin() + -->
<!--   scale_fill_brewer(palette = "Accent") + -->
<!--   facet_wrap(~type, scales = "free_y") + -->
<!--   labs(x = "min. dist. (km)", y = "raw material", fill = "Region") + -->
<!--   theme_minimal() + -->
<!--   theme(legend.position = "bottom",  -->
<!--         panel.border = element_rect(color = "black", fill = NA)) -->
<!-- ``` -->

@@

```{r dem}
dem_terra <- terra::rast(here::here(derived_data, "/dem/dem_aster_sjtsk_stars.tif"))
```

@@

<!--
```{r dem-data}
dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))

# bind to results
res1_long <- set_base$period1 %>% 
prep_dem_res() %>% 
bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
select(-facet) %>% 
prep_dem_res() %>% 
bind_rows(res2_long)
```

```{r altitude-to-res}
altitude <- set_spat %>% st_drop_geometry() %>% 
as_tibble() %>% 
select(id, altitude)

# bind to results
res1_long <- set_base$period1 %>% 
prep_alt_res() %>% 
bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
select(-facet) %>% 
prep_alt_res() %>% 
bind_rows(res2_long)

```

```{r dem-analysis-correlation, fig.cap='Correlation between altitude derived from DEM and altitude reading in the topographic map.'}
alt_plot <- set_spat %>% st_drop_geometry() %>% 
as_tibble() %>% 
select(id, altitude) %>% 
left_join(dem_data)

alt_cor <- alt_plot %>% 
summarise(cor = cor(altitude, dem)) %>% 
pull(cor) %>% 
round(2)

alt_plot %>% 
ggplot(aes(altitude, dem)) +
geom_point() +
theme_minimal() +
annotate(geom = "text", x = 175, y = 525, 
label = paste("Cor. ", alt_cor))
```
-->

@@

```{r aspect-norm}
shapiro.test(filter(med_asp, reg == "B")$med_asp)
shapiro.test(filter(med_asp, reg == "M")$med_asp)
```

```{r aspect-plot-reg, fig.cap="Histogram of aspect distibution on polar coordinates."}

# polar plot
med_asp %>% 
  ggplot(aes(med_asp)) + 
  geom_histogram(color = "black", fill = "white") +
  scale_x_continuous(limits = c(0, 359), 
                     breaks = c(0, 90, 180, 270)) +
  coord_polar(start = 0) + 
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(x = "median aspect (°)")
```

```{r aspect-plot-reg-p2}
med_asp %>% 
  inner_join(settlements2, by = "id") %>% 
  ggplot(aes(med_asp)) + 
  geom_histogram(color = "black", fill = "white") +
  scale_x_continuous(limits = c(0, 360), 
                     breaks = c(0, 90, 180, 270, 360)) +
  facet_grid(vars(period_label), vars(reg), scales = "free_y") +
  theme_bw() +
  labs(x = "median aspect (°)")
```

```{r aspect-clean}
rm(list = c("dem_asp", "asp", "med_asp"))
```

@@

```{r tpi}
dem_tpi <- terra::terrain(dem_terra, v = "TPI")

# tpi in the vicinity of settlements - 2km
tpi <- terra::extract(dem_tpi, 
                      terra::vect(st_buffer(set_spat, dist = 2e3)))

@@

<!--
# Settlement 'continuity'

@@

res2_long %>% 
plot_continuity(y = n_set_per2)
```
-->

@@

```{r res-violin-out}
ggsave(here::here("analysis/figures", "res1_violin.png"), res1_violin, 
       width = 9, height = 12)
ggsave(here::here("analysis/figures", "res2_violin.png"), res2_violin, 
       width = 9, height = 12)
```

@@

```{r res-line-out}
ggsave(here::here("analysis/figures", "res1_line.png"), res1_line, 
       width = 7, height = 7)
ggsave(here::here("analysis/figures", "res2_line.png"), res2_line, 
       width = 7, height = 7)
```

@@

<!-- ```{r pca1-all} -->
<!-- # fit PCA -->
<!-- pca1_all_fit <- res1_wide %>%  -->
<!--   # select(-cont) %>%  -->
<!--   select(where(is.numeric)) %>%  -->
<!--   scale(center = TRUE, scale = TRUE) %>%  -->
<!--   prcomp() -->

<!-- # augmented fitted values -->
<!-- pca1_all_aug <- pca1_all_fit %>%  -->
<!--   broom::augment(res1_wide) %>%  -->
<!--   select(-".rownames") %>%  -->
<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
<!--   settlements::add_region() %>%  -->
<!--   settlements::add_period_label() %>%  -->
<!--   settlements::add_neo() -->

<!-- # rotation matrix -->
<!-- pca1_all_rot <- pca1_all_fit %>% -->
<!--   broom::tidy(matrix = "rotation") %>% -->
<!--   mutate(PC = str_c("PC", PC)) %>% -->
<!--   pivot_wider(names_from = PC) -->

<!-- # var explained -->
<!-- pca1_all_var_exp <- pca1_all_fit %>%  -->
<!--   broom::augment(res1_wide) %>% -->
<!--   summarise(across(contains("PC"), var)) %>%  -->
<!--   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%  -->
<!--   mutate(var_exp = variance/sum(variance), -->
<!--          cum_var_exp = cumsum(var_exp), -->
<!--          pc = str_replace(pc, ".fitted", "")) %>%  -->
<!--   select(pc, var_exp, cum_var_exp) -->

<!-- # PC cum var explained over 0.9 -->
<!-- pca1_all_var_exp_over <- pca1_all_var_exp %>%  -->
<!--   mutate(n = !(cum_var_exp >= 0.9)) %>%  -->
<!--   pull(n) %>%  -->
<!--   sum() + 1 -->
<!-- ``` -->

<!-- ```{r pca1-all-var-exp-tab} -->
<!-- knitr::kable(pca1_all_var_exp, col.names = c("Comp.", -->
<!--                                              "Prop. of variance",  -->
<!--                                              "Cumulative prop.")) -->
<!-- ``` -->

<!-- ```{r pca1-all-loadings-plot} -->
<!-- # PCA loadings -->
<!-- pca1_all_loadings <- pca1_all_fit %>%  -->
<!--   broom::tidy(matrix = "rotation") %>%  -->
<!--   mutate(component = str_c("PC", PC), -->
<!--          component = fct_reorder(component, PC)) %>%  -->
<!--   filter(PC <= pca1_all_var_exp_over) %>%  -->
<!--   group_by(PC) %>%  -->
<!--   top_n(8, abs(value)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(cor = if_else(value > 0, "positive", "negative"), -->
<!--          column = tidytext::reorder_within(column, abs(value), component)) -->

<!-- pca1_all_loadings %>%  -->
<!--   ggplot(aes(abs(value), column, fill = cor)) + -->
<!--   geom_col(color = "black") + -->
<!--   scale_fill_manual(values = c("white", "gray80")) + -->
<!--   facet_wrap(vars(component), scales = "free_y") + -->
<!--   tidytext::scale_y_reordered() + -->
<!--   theme_bw() + -->
<!--   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") -->
<!-- ``` -->

<!-- ```{r plot-pca1-results} -->
<!-- plot_pca1_full <- function(x = "PC1", y = "PC2") { -->

<!--   pcs <- c(x, y) -->

<!--   pca1_all_plot <- pca1_all_aug %>%  -->
<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
<!--     geom_point(data = select(pca1_all_aug, -period_label),  -->
<!--                alpha = 0.05, size = 0.4) + -->
<!--     geom_point(aes(color = reg),  -->
<!--                alpha = 0.8, size = 0.5, show.legend = FALSE) + -->
<!--     scale_color_manual(values = mycol[c(1, 2)]) + -->
<!--     facet_grid(vars(period_label), vars(reg)) + -->
<!--     theme_minimal() -->

<!--   pca1_all_rot <-  pca1_all_rot %>% -->
<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
<!--     geom_segment(xend = 0, yend = 0, alpha = 0.4) + -->
<!--     geom_text( -->
<!--       aes(label = column), -->
<!--       hjust = 1, nudge_x = -0.02, -->
<!--       color = "#904C2F", -->
<!--       size = 2.6 -->
<!--     ) + -->
<!--     theme_minimal() -->

<!--   pca1_all_load <- pca1_all_loadings %>%  -->
<!--     filter(component %in% pcs) %>%  -->
<!--     ggplot(aes(abs(value), column, fill = cor)) + -->
<!--     geom_col(color = "black", show.legend = FALSE) + -->
<!--     scale_fill_manual(values = c("white", "gray80")) + -->
<!--     facet_wrap(vars(component), scales = "free_y") + -->
<!--     tidytext::scale_y_reordered() + -->
<!--     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") + -->
<!--     theme_minimal() -->

<!--   pca1_all_plot / (pca1_all_rot | pca1_all_load) +  -->
<!--     plot_layout(heights = c(6, 1)) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc1-pc2} -->
<!-- plot_pca1_full() -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc12.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc3-pc4} -->
<!-- plot_pca1_full("PC3", "PC4") -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc34.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc5-pc6} -->
<!-- plot_pca1_full("PC5", "PC6") -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc56.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc7-pc8} -->
<!-- plot_pca1_full("PC7", "PC8") -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc78.png"), height = 14) -->
<!-- ``` -->

<!-- ### Pottery groups -->

<!-- ```{r pca2-all} -->
<!-- # fit PCA -->
<!-- pca2_all_fit <- res2_wide %>%  -->
<!--   # select(-cont) %>%  -->
<!--   select(where(is.numeric)) %>%  -->
<!--   scale(center = TRUE, scale = TRUE) %>%  -->
<!--   prcomp() -->

<!-- # augmented fitted values -->
<!-- pca2_all_aug <- pca2_all_fit %>%  -->
<!--   broom::augment(res2_wide) %>%  -->
<!--   select(-".rownames") %>%  -->
<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
<!--   settlements::add_region() %>%  -->
<!--   settlements::add_period_label() %>%  -->
<!--   settlements::add_neo() -->

<!-- # rotation matrix -->
<!-- pca2_all_rot <- pca2_all_fit %>% -->
<!--   broom::tidy(matrix = "rotation") %>% -->
<!--   mutate(PC = str_c("PC", PC)) %>% -->
<!--   pivot_wider(names_from = PC) -->

<!-- # var explained -->
<!-- pca2_all_var_exp <- pca2_all_fit %>%  -->
<!--   broom::augment(res2_wide) %>% -->
<!--   summarise(across(contains("PC"), var)) %>%  -->
<!--   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%  -->
<!--   mutate(var_exp = variance/sum(variance), -->
<!--          cum_var_exp = cumsum(var_exp), -->
<!--          pc = str_replace(pc, ".fitted", "")) %>%  -->
<!--   select(pc, var_exp, cum_var_exp) -->

<!-- # PC cum var explained over 0.9 -->
<!-- pca2_all_var_exp_over <- pca2_all_var_exp %>%  -->
<!--   mutate(n = !(cum_var_exp >= 0.9)) %>%  -->
<!--   pull(n) %>%  -->
<!--   sum() + 1 -->
<!-- ``` -->

<!-- <!-- -->
<!-- ```{r} -->
<!-- fuu <- res2_wide %>% mutate(across(where(is.numeric), \(x) x == 0)) %>%  -->
<!-- select(where(is.logical)) %>%  -->
<!-- rowSums() %>%  -->
<!-- as_tibble() -->

<!-- shit <- res2_wide %>% select(id, period) %>% bind_cols(fuu) %>% arrange(desc(value), id, period) %>%  -->
<!-- filter(value == 8) %>% pull(id) %>% factor() %>% levels() -->

<!-- settlements2 %>%  -->
<!-- filter(id %in% shit) %>%  -->
<!-- sf::st_transform(4326) %>%  -->
<!-- leaflet::leaflet() %>%  -->
<!-- leaflet::addTiles() %>%  -->
<!-- leaflet::addMarkers() -->

<!-- res2_wide %>% filter(id %in% shit) %>% View() -->
<!-- ``` -->
<!-- -->

<!-- ```{r pca2-all-var-exp-tab} -->
<!-- knitr::kable(pca1_all_var_exp, col.names = c("Comp.", -->
<!--                                              "Prop. of variance",  -->
<!--                                              "Cumulative prop.")) -->
<!-- ``` -->

<!-- ```{r pca2-all-loadings-plot} -->
<!-- # PCA loadings -->
<!-- pca2_all_loadings <- pca2_all_fit %>%  -->
<!--   broom::tidy(matrix = "rotation") %>%  -->
<!--   mutate(component = str_c("PC", PC), -->
<!--          component = fct_reorder(component, PC)) %>%  -->
<!--   filter(PC <= pca2_all_var_exp_over) %>%  -->
<!--   group_by(PC) %>%  -->
<!--   top_n(8, abs(value)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(cor = if_else(value > 0, "positive", "negative"), -->
<!--          column = tidytext::reorder_within(column, abs(value), component)) -->

<!-- pca2_all_loadings %>%  -->
<!--   ggplot(aes(abs(value), column, fill = cor)) + -->
<!--   geom_col(color = "black") + -->
<!--   scale_fill_manual(values = c("white", "gray80")) + -->
<!--   facet_wrap(vars(component), scales = "free_y") + -->
<!--   tidytext::scale_y_reordered() + -->
<!--   theme_bw() + -->
<!--   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") -->
<!-- ``` -->

<!-- ```{r plot-pca2-results} -->
<!-- plot_pca2_full <- function(x = "PC1", y = "PC2") { -->

<!--   pcs <- c(x, y) -->

<!--   pca2_all_plot <- pca2_all_aug %>%  -->
<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
<!--     geom_point(data = select(pca2_all_aug, -period_label),  -->
<!--                alpha = 0.05, size = 0.4) + -->
<!--     geom_point(aes(color = reg),  -->
<!--                alpha = 0.8, size = 0.5, show.legend = FALSE) + -->
<!--     scale_color_manual(values = mycol[c(1, 2)]) + -->
<!--     facet_grid(vars(period_label), vars(reg)) + -->
<!--     theme_minimal() -->

<!--   pca2_all_rot <-  pca2_all_rot %>% -->
<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
<!--     geom_segment(xend = 0, yend = 0, alpha = 0.4) + -->
<!--     geom_text( -->
<!--       aes(label = column), -->
<!--       hjust = 1, nudge_x = -0.02, -->
<!--       color = "#904C2F", -->
<!--       size = 2.6 -->
<!--     ) + -->
<!--     theme_minimal() -->

<!--   pca2_all_load <- pca2_all_loadings %>%  -->
<!--     filter(component %in% pcs) %>%  -->
<!--     ggplot(aes(abs(value), column, fill = cor)) + -->
<!--     geom_col(color = "black", show.legend = FALSE) + -->
<!--     scale_fill_manual(values = c("white", "gray80")) + -->
<!--     facet_wrap(vars(component), scales = "free_y") + -->
<!--     tidytext::scale_y_reordered() + -->
<!--     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") + -->
<!--     theme_minimal() -->

<!--   pca2_all_plot / (pca2_all_rot | pca2_all_load) +  -->
<!--     plot_layout(heights = c(6, 1)) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r pca2-all-plots-pc1-pc2} -->
<!-- plot_pca2_full() -->
<!-- ggsave(here::here("analysis/figures/pca2_full_pc12.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca2-all-plots-pc3-pc4} -->
<!-- plot_pca2_full("PC3", "PC4") -->
<!-- ggsave(here::here("analysis/figures/pca2_full_pc34.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca2-all-plots-pc5-pc6} -->
<!-- plot_pca2_full("PC5", "PC6") -->
<!-- ggsave(here::here("analysis/figures/pca2_full_pc56.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca2-all-plots-pc7-pc8} -->
<!-- plot_pca2_full("PC7", "PC8") -->
<!-- ggsave(here::here("analysis/figures/pca2_full_pc78.png"), height = 14) -->
<!-- ``` -->

@@

ggsave(here::here("analysis/paper/temp/pca_skup_1.png"), width = 10, height = 21)
```

@@

ggsave(here::here("analysis/paper/temp/pca_skup_2.png"), width = 10, height = 21)
```

@@

## Groups in the data

This is not done for pottery traditions due to an error in one of the periods/
regions that is caused by not enough variance in the PCA data. Weird...

```{r package-mclust}
library(mclust)
```

```{r mclust2-neo}
mb_clust2 <- pca2_neo %>% mbc_cluster()

x <- mb_clust2 %>%
  dplyr::select(reg, neo, classif) %>% 
  tidyr::unnest() %>% 
  dplyr::ungroup(c(reg, neo)) %>% 
  dplyr::mutate(g = str_c(reg, g))

set_spat %>% 
  inner_join(x, by = "id") %>% 
  ggplot() +
  geom_sf(aes(color = g)) +
  facet_grid(vars(g), vars(neo))

```

mb_classif2_neoB <- mb_clust2 %>% mbc_classify(neo = "NeoB")
mb_classif2_neoC <- mb_clust2 %>% mbc_classify(neo = "NeoC")

(mb_classif2_neoB == mb_classif2_neoC) %>% all()

>>>>

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  # fig.path = "../figures/",
  dpi = 300
)
```

@@

```{r get-data, cache = FALSE}
derived_data <- "analysis/data/derived_data"
input_data <- "analysis/data/input_data"
path_maps <- here::here(derived_data, "maps/")
path_data <- here::here(derived_data)
path_temp <- here::here(derived_data, "temp/")

@@

# labels
labs_chrono <- read_rds(here::here(input_data, "chrono_labels.RDS"))

# database
set_base <- read_rds(here::here(input_data, "settlements.RDS"))

set_spat <- st_read(here::here(input_data, "settlements_sf.geojson"), 
                    quiet = TRUE)

@@

```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."}
# rm_dist %>%
#   ggplot(aes(x = min, y = rm, fill = region)) +
#   geom_violin() +
#   scale_fill_brewer(palette = "Accent") +
#   facet_wrap(~type, scales = "free_y") +
#   labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
#   theme_minimal() +
#   theme(legend.position = "bottom",
#         panel.border = element_rect(color = "black", fill = NA))
```

@@

```{r dem}
dem_terra <- terra::rast(here::here(derived_data, "archive/dem/dem_aster_sjtsk_stars.tif"))
```

@@

```{r dem-data}
# dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))
# 
# # bind to results
# res1_long <- set_base$period1 %>% 
# prep_dem_res() %>% 
# bind_rows(res1_long)
# 
# res2_long <- set_base$period2 %>% 
# select(-facet) %>% 
# prep_dem_res() %>% 
# bind_rows(res2_long)
```

```{r altitude-to-res}
# altitude <- set_spat %>% st_drop_geometry() %>% 
# as_tibble() %>% 
# select(id, altitude)
# 
# # bind to results
# res1_long <- set_base$period1 %>% 
# prep_alt_res() %>% 
# bind_rows(res1_long)
# 
# res2_long <- set_base$period2 %>% 
# select(-facet) %>% 
# prep_alt_res() %>% 
# bind_rows(res2_long)
```

```{r dem-analysis-correlation, fig.cap='Correlation between altitude derived from DEM and altitude reading in the topographic map.'}
# alt_plot <- set_spat %>% st_drop_geometry() %>% 
# as_tibble() %>% 
# select(id, altitude) %>% 
# left_join(dem_data)
# 
# alt_cor <- alt_plot %>% 
# summarise(cor = cor(altitude, dem)) %>% 
# pull(cor) %>% 
# round(2)
# 
# alt_plot %>% 
# ggplot(aes(altitude, dem)) +
# geom_point() +
# theme_minimal() +
# annotate(geom = "text", x = 175, y = 525, 
# label = paste("Cor. ", alt_cor))
```

@@

```{r aspect-norm}
# shapiro.test(filter(med_asp, reg == "B")$med_asp)
# shapiro.test(filter(med_asp, reg == "M")$med_asp)
```

```{r aspect-plot-reg, fig.cap="Histogram of aspect distibution on polar coordinates."}
# # polar plot
# med_asp %>% 
#   ggplot(aes(med_asp)) + 
#   geom_histogram(color = "black", fill = "white") +
#   scale_x_continuous(limits = c(0, 359), 
#                      breaks = c(0, 90, 180, 270)) +
#   coord_polar(start = 0) + 
#   facet_wrap(vars(reg)) +
#   theme_bw() +
#   labs(x = "median aspect (°)")
```

```{r aspect-plot-reg-p2}
# med_asp %>% 
#   inner_join(settlements2, by = "id") %>% 
#   ggplot(aes(med_asp)) + 
#   geom_histogram(color = "black", fill = "white") +
#   scale_x_continuous(limits = c(0, 360), 
#                      breaks = c(0, 90, 180, 270, 360)) +
#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
#   theme_bw() +
#   labs(x = "median aspect (°)")
```

```{r aspect-clean2}
rm(list = c("dem_asp", "asp", "med_asp"))
```

@@

```{r tpi}
# dem_tpi <- terra::terrain(dem_terra, v = "TPI")
# 
# # tpi in the vicinity of settlements - 2km
# tpi <- terra::extract(dem_tpi, 
#                       terra::vect(st_buffer(set_spat, dist = 2e3)))
# 
# # derive median aspect
# mean_slope <- slope %>% 
#   tibble() %>% 
#   group_by(ID) %>% 
#   summarise(mean_slope = mean(slope)) %>% 
#   bind_cols(id = set_spat$id) %>% 
#   mutate(reg = str_extract(id, "^."))

@@

# Settlement 'continuity'

@@

res2_long %>% 
plot_continuity(y = n_set_per2)
```

@@

```{r res-violin-out}
# ggsave(here::here("analysis/figures", "res1_violin.png"), res1_violin, 
#        width = 9, height = 12)
# ggsave(here::here("analysis/figures", "res2_violin.png"), res2_violin, 
#        width = 9, height = 12)
```

@@

```{r res-line-out}
# ggsave(here::here("analysis/figures", "res1_line.png"), res1_line, 
#        width = 7, height = 7)
# ggsave(here::here("analysis/figures", "res2_line.png"), res2_line, 
#        width = 7, height = 7)
```

@@

```{r pca1-all}
# # fit PCA
# pca1_all_fit <- res1_wide %>%
#   # select(-cont) %>%
#   select(where(is.numeric)) %>%
#   scale(center = TRUE, scale = TRUE) %>%
#   prcomp()
# 
# # augmented fitted values
# pca1_all_aug <- pca1_all_fit %>%
#   broom::augment(res1_wide) %>%
#   select(-".rownames") %>%
#   rename_with(\(x) str_remove(x, ".fitted")) %>%
#   settlements::add_region() %>%
#   settlements::add_period_label() %>%
#   settlements::add_neo()
# 
# # rotation matrix
# pca1_all_rot <- pca1_all_fit %>%
#   broom::tidy(matrix = "rotation") %>%
#   mutate(PC = str_c("PC", PC)) %>%
#   pivot_wider(names_from = PC)
# 
# # var explained
# pca1_all_var_exp <- pca1_all_fit %>%
#   broom::augment(res1_wide) %>%
#   summarise(across(contains("PC"), var)) %>%
#   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%
#   mutate(var_exp = variance/sum(variance),
#          cum_var_exp = cumsum(var_exp),
#          pc = str_replace(pc, ".fitted", "")) %>%
#   select(pc, var_exp, cum_var_exp)
# 
# # PC cum var explained over 0.9
# pca1_all_var_exp_over <- pca1_all_var_exp %>%
#   mutate(n = !(cum_var_exp >= 0.9)) %>%
#   pull(n) %>%
#   sum() + 1
```

```{r pca1-all-var-exp-tab}
# knitr::kable(pca1_all_var_exp, col.names = c("Comp.",
#                                              "Prop. of variance",
#                                              "Cumulative prop."))
```

```{r pca1-all-loadings-plot}
# # PCA loadings
# pca1_all_loadings <- pca1_all_fit %>%
#   broom::tidy(matrix = "rotation") %>%
#   mutate(component = str_c("PC", PC),
#          component = fct_reorder(component, PC)) %>%
#   filter(PC <= pca1_all_var_exp_over) %>%
#   group_by(PC) %>%
#   top_n(8, abs(value)) %>%
#   ungroup() %>%
#   mutate(cor = if_else(value > 0, "positive", "negative"),
#          column = tidytext::reorder_within(column, abs(value), component))
# 
# pca1_all_loadings %>%
#   ggplot(aes(abs(value), column, fill = cor)) +
#   geom_col(color = "black") +
#   scale_fill_manual(values = c("white", "gray80")) +
#   facet_wrap(vars(component), scales = "free_y") +
#   tidytext::scale_y_reordered() +
#   theme_bw() +
#   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value")
```

```{r plot-pca1-results}
# plot_pca1_full <- function(x = "PC1", y = "PC2") {
# 
#   pcs <- c(x, y)
# 
#   pca1_all_plot <- pca1_all_aug %>%
#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
#     geom_point(data = select(pca1_all_aug, -period_label),
#                alpha = 0.05, size = 0.4) +
#     geom_point(aes(color = reg),
#                alpha = 0.8, size = 0.5, show.legend = FALSE) +
#     scale_color_manual(values = mycol[c(1, 2)]) +
#     facet_grid(vars(period_label), vars(reg)) +
#     theme_minimal()
# 
#   pca1_all_rot <-  pca1_all_rot %>%
#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
#     geom_segment(xend = 0, yend = 0, alpha = 0.4) +
#     geom_text(
#       aes(label = column),
#       hjust = 1, nudge_x = -0.02,
#       color = "#904C2F",
#       size = 2.6
#     ) +
#     theme_minimal()
# 
#   pca1_all_load <- pca1_all_loadings %>%
#     filter(component %in% pcs) %>%
#     ggplot(aes(abs(value), column, fill = cor)) +
#     geom_col(color = "black", show.legend = FALSE) +
#     scale_fill_manual(values = c("white", "gray80")) +
#     facet_wrap(vars(component), scales = "free_y") +
#     tidytext::scale_y_reordered() +
#     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") +
#     theme_minimal()
# 
#   pca1_all_plot / (pca1_all_rot | pca1_all_load) +
#     plot_layout(heights = c(6, 1))
# }
```

```{r pca1-all-plots-pc1-pc2}
# plot_pca1_full()
# ggsave(here::here("analysis/figures/pca1_full_pc12.png"), height = 14)
```

```{r pca1-all-plots-pc3-pc4}
# plot_pca1_full("PC3", "PC4")
# ggsave(here::here("analysis/figures/pca1_full_pc34.png"), height = 14)
```

```{r pca1-all-plots-pc5-pc6}
# plot_pca1_full("PC5", "PC6")
# ggsave(here::here("analysis/figures/pca1_full_pc56.png"), height = 14)
```

```{r pca1-all-plots-pc7-pc8}
# plot_pca1_full("PC7", "PC8")
# ggsave(here::here("analysis/figures/pca1_full_pc78.png"), height = 14)
```

```{r pca2-all}
# # fit PCA
# pca2_all_fit <- res2_wide %>%
#   # select(-cont) %>%
#   select(where(is.numeric)) %>%
#   scale(center = TRUE, scale = TRUE) %>%
#   prcomp()
# 
# # augmented fitted values
# pca2_all_aug <- pca2_all_fit %>%
#   broom::augment(res2_wide) %>%
#   select(-".rownames") %>%
#   rename_with(\(x) str_remove(x, ".fitted")) %>%
#   settlements::add_region() %>%
#   settlements::add_period_label() %>%
#   settlements::add_neo()
# 
# # rotation matrix
# pca2_all_rot <- pca2_all_fit %>%
#   broom::tidy(matrix = "rotation") %>%
#   mutate(PC = str_c("PC", PC)) %>%
#   pivot_wider(names_from = PC)
# 
# # var explained
# pca2_all_var_exp <- pca2_all_fit %>%
#   broom::augment(res2_wide) %>%
#   summarise(across(contains("PC"), var)) %>%
#   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%
#   mutate(var_exp = variance/sum(variance),
#          cum_var_exp = cumsum(var_exp),
#          pc = str_replace(pc, ".fitted", "")) %>%
#   select(pc, var_exp, cum_var_exp)
# 
# # PC cum var explained over 0.9
# pca2_all_var_exp_over <- pca2_all_var_exp %>%
#   mutate(n = !(cum_var_exp >= 0.9)) %>%
#   pull(n) %>%
#   sum() + 1
```

```{r pca2-all-var-exp-tab}
# knitr::kable(pca1_all_var_exp, col.names = c("Comp.",
#                                              "Prop. of variance",
#                                              "Cumulative prop."))
```

```{r pca2-all-loadings-plot}
# # PCA loadings
# pca2_all_loadings <- pca2_all_fit %>%
#   broom::tidy(matrix = "rotation") %>%
#   mutate(component = str_c("PC", PC),
#          component = fct_reorder(component, PC)) %>%
#   filter(PC <= pca2_all_var_exp_over) %>%
#   group_by(PC) %>%
#   top_n(8, abs(value)) %>%
#   ungroup() %>%
#   mutate(cor = if_else(value > 0, "positive", "negative"),
#          column = tidytext::reorder_within(column, abs(value), component))
# 
# pca2_all_loadings %>%
#   ggplot(aes(abs(value), column, fill = cor)) +
#   geom_col(color = "black") +
#   scale_fill_manual(values = c("white", "gray80")) +
#   facet_wrap(vars(component), scales = "free_y") +
#   tidytext::scale_y_reordered() +
#   theme_bw() +
#   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value")
```

```{r plot-pca2-results}
# plot_pca2_full <- function(x = "PC1", y = "PC2") {
# 
#   pcs <- c(x, y)
# 
#   pca2_all_plot <- pca2_all_aug %>%
#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
#     geom_point(data = select(pca2_all_aug, -period_label),
#                alpha = 0.05, size = 0.4) +
#     geom_point(aes(color = reg),
#                alpha = 0.8, size = 0.5, show.legend = FALSE) +
#     scale_color_manual(values = mycol[c(1, 2)]) +
#     facet_grid(vars(period_label), vars(reg)) +
#     theme_minimal()
# 
#   pca2_all_rot <-  pca2_all_rot %>%
#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
#     geom_segment(xend = 0, yend = 0, alpha = 0.4) +
#     geom_text(
#       aes(label = column),
#       hjust = 1, nudge_x = -0.02,
#       color = "#904C2F",
#       size = 2.6
#     ) +
#     theme_minimal()
# 
#   pca2_all_load <- pca2_all_loadings %>%
#     filter(component %in% pcs) %>%
#     ggplot(aes(abs(value), column, fill = cor)) +
#     geom_col(color = "black", show.legend = FALSE) +
#     scale_fill_manual(values = c("white", "gray80")) +
#     facet_wrap(vars(component), scales = "free_y") +
#     tidytext::scale_y_reordered() +
#     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") +
#     theme_minimal()
# 
#   pca2_all_plot / (pca2_all_rot | pca2_all_load) +
#     plot_layout(heights = c(6, 1))
# }
```

```{r pca2-all-plots-pc1-pc2}
# plot_pca2_full()
# ggsave(here::here("analysis/figures/pca2_full_pc12.png"), height = 14)
```

```{r pca2-all-plots-pc3-pc4}
# plot_pca2_full("PC3", "PC4")
# ggsave(here::here("analysis/figures/pca2_full_pc34.png"), height = 14)
```

```{r pca2-all-plots-pc5-pc6}
# plot_pca2_full("PC5", "PC6")
# ggsave(here::here("analysis/figures/pca2_full_pc56.png"), height = 14)
```

```{r pca2-all-plots-pc7-pc8}
# plot_pca2_full("PC7", "PC8")
# ggsave(here::here("analysis/figures/pca2_full_pc78.png"), height = 14)
```

@@

# ggsave(here::here("analysis/paper/temp/pca_skup_1.png"), width = 10, height = 21)
```

@@

# ggsave(here::here("analysis/paper/temp/pca_skup_2.png"), width = 10, height = 21)
```

@@
